```{r variables-storing}
 # generate a random sample of 20 data points and store it in a variable called random_sample
random_sample <- sample(1:100, 20, replace =TRUE)

# call random_sample which outputs the values to the screen
random_sample

# calculate the mean of random_sample and store it in a variable called random_sample_mean
random_sample_mean <- mean(random_sample) 

# call random_sample_mean which outputs the values to the screen
random_sample_mean 
```

<div class = 'exercise'>
Let's build our own function that calculates a mean to see what's all happening under the hood.

First we define the mean as the sum of all data points divided by the number of data points. We'll leverage two existing functions, `sum()` and `length()`, the former of which performs simple addition and the latter of which tells us how many data points are in a vector.
</div>

<div class = 'solution'>
```{r mean-function, class.source = 'fold-hide', results = FALSE}
# define our function. The function takes one input, which is a data source, it then add all the data points together and divides it by the number of data points.
# the function is stored in a variable called my_mean
my_mean <- function(data) {
  total <- sum(data) # add the data
  n = length(data) # count the data
  calculation = total/n # do the math
  calculation # print the output
}

my_data <- sample(1:100, 20, replace = F) # generate some data

my_mean(my_data) # calculate the mean using my_mean
```
</div>

<div class = 'summary'>
Functions take in data and do things with that data. Variables allow us to store our data for easy access.
</div>
