---
Title: "R For Beginners"
pagetitle: "R For Beginners"
output:
  html_document:
    includes:
      in_header: header.html
---
```{r echo=FALSE}
knitr::opts_chunk$set(comment = NA)
knitr::read_chunk("./scripts/workshop_files.R")
```

This workshop is offered through UBC Okanagan's [Centre for Scholarly Communications](https://csc.ok.ubc.ca). Questions can be directed to [csc.ok@ubc.ca](mailto:csc.ok@ubc.ca).

## Description

This four-part workshop is designed for the absolute beginner to R with no programming experience in R or other languages.

**Session 1** will cover what R is and introduce users to a common application for programming in R, RStudio. We'll cover basic concepts related to conducting academic work with a programming language, write a simple script, and explore options for getting help.

**Session 2** will get a little deeper into how R works, beginning with an introduction to R's built in data types and structures. From there we'll start importing data and review best practices for handling data imports.

**Session 3** will then dive into getting a data set ready for subsequent analysis. This will involve creating new variables from existing variables, working with categorical data, an overview of dates, and basic sub-setting to isolate observations and variables.

**Session 4** will cover adding increased functionality to your programming environment with additional packages and building plots using the popular package ggplot2.

You will need a laptop for this workshop. You will also need to install R and RStudio before coming to the first session. Instructions for getting these set up will be provided after registration.

## Outcomes

This is an introductory workshop intended for individuals with little to no familiarity with R or similar tools that rely on a scripted environment. It is intended as a low barrier entry into the world of using your computer to engage with data outside of graphical tools like Excel or SPSS. We will not cover statistical analyses.

Instead, the hope is that you will walk away with a base set of skills and language to engage in some self directed learning that might involve more workshops on more advanced applications of R, texts available through the library or on-line, and forums where other users are asking questions and sharing their solutions.

## Schedule

| Date | Time | Location |
| :- | :- | :- |
| January 25, 2023 | 2:00pm - 3:00pm | LIB 111 |
| February 1, 2023 | 2:00pm - 3:00pm | LIB 111 |
| February 8, 2023 | 2:00pm - 3:00pm | LIB 111 |
| February 15, 2023 | 2:00pm - 3:00pm | LIB 111 |

### Files

Right click or ctrl + left click on a Mac to save files.

```{r loadfiles, results='asis', echo=FALSE}
```
