---
title: Subsetting
output:
  html_document:
    code_folding: show
---

Another common activity is subsetting data -- isolating a portion of the data, usually based on a condition. There are a few different ways we can subset in R, one of which we've already kind of seen with indexing, we just didn't assign the output to a variable.

```{r}
sq_data_age <- sq_data$Age # select one column

sq_data_sounds <- sq_data[,22:24] # select multiple columns
```

<div class = 'exercise'>
Using the above as a model, take a random sample of ten observations (rows) from `sq_data`.
</div>

<div class = 'solution'>
```{r, class.source = 'fold-hide', results = FALSE}
# if we know the number of rows
sq_data_random_sample <- sq_data[sample(1:3023, 10),]

# or, better yet, if the number of rows might be variable

sq_data_random_sample <- sq_data[sample(1:length(sq_data), 10),]
```
</div>

Subsetting on conditions can be achieved with `subset()`. We'll provide `subset()` with three arguments:

1. A data set to subset from
2. The condition on which to subset
3. The variables from the original data set to keep -- leave this empty to default to all original variables

<div class = "note">
R allows us to specify several conditions

* less than `>`
* greater than `<`
* less than or equal to `<=` 
* greater than or equal `>=`
* equivalent to `==`
* not equivalent to `!=`
* or `|`

Note that `=` is equivalent to `<-`, setting a value, while `==` tests whether or not two things are the same.

```{r}
(x = 2)
(x == 2)
```
</div>

Let's say we only want to see the records for adult squirrels.

```{r}
sq_data_adult <- subset(sq_data, Age == "Adult") # from sq_data, select only rows where Age equals Adult
```

Or all records with cinnamon coloured adult squirrels.

```{r}
sq_data_c_adult <- subset(sq_data, Age == "Adult" & Primary.Fur.Color == "Cinnamon")
```

Or all records where an age has been recorded. This can be done in two ways

```{r}
sq_data_age_1 <- subset(sq_data, !is.na(Age))

sq_data_age_2 <- subset(sq_data, Age == "Adult" | Age == "Juvenile")
```

<div class = "note">
Remember, `NA` values are a special value, indicating that a value is missing. This is different from writing the characters NA. `is.na()` tests for `NA`. It's for this reason that running `subset(sq_data, sq_data$Age == 'NA')` will give you an empty data frame.
</div>

Or maybe we want to investigate patterns of only those squirrels for which we don't have age recorded!

```{r}
sq_data_age_na <- subset(sq_data, is.na(Age))
```

Lastly, before an exercise, we might want only observations that are complete, that is, records with no missing values

```{r}
sq_data_complete <- sq_data[complete.cases(sq_data), ] # not terribly useful with this particular data set
```

<div class = 'exercise'>
All data was collected between October 6 and October 20, 2018.  Create a subset of the data that includes only data collected on or before October 14, and only retains the variables `Date` and `Age`. Verify your date range output with `unique()` and if you get that far, order this output from `unique()` using `sort()` for easier reading.

Remember that we have not yet converted our `Date` variable to a date class. 
</div>

<div class = 'solution'>
```{r, class.source = 'fold-hide', results = FALSE}
sq_data$Date <- as.Date(sq_data$Date, "%m%d%Y")
sq_data_date <- subset(sq_data, Date <= "2018-10-14", select = c(Date, Age))
sort(unique(sq_data_date$Date))
```
</div>

<div class = 'summary'>
Subsetting can be achieved in many ways. We can use indexing to get specific observations (rows) and / or variable (columns), we can use other functions like `is.na()` and `complete.cases()` to work around missing values, and we can use `subset()` to conditionally subset.
</div>

<div class = 'functions'>
| Function | Description |
| :--- | :--- |
| `subset` | conditionally susbet a data frame |
| `is.na` | conditionally test if values in a vector are `NA`. |
| `complete.cases` | conditionally test if cases (rows) are complete. |
</div>