---
title: "Functions"
output:
  html_document:
    code_folding: show
---

```{r echo=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      comment = NA)
```

Functions take in data and do things with that data. We can write our own functions, but generally in R we'll be using functions that have already been written or built. You've already seen three functions, `sqrt()`, `log10`, and `log2`.

Calling a function requires 2 things: the function itself and any arguments the function allows us to specify -- one of these arguments is the data source, but there are usually other parameters we can specify.

## Rivers & Descriptive Stats

We'll explore the basics of a handful of functions for descriptive stats with a built in data set called rivers. We can look at the data in rivers simply by calling it:

```{r}
rivers # call the rivers dataset
```
rivers consists of a single variable -- lengths in miles of a set of rivers in the United States.

To calculate the mean of these data, we use the function `mean()`

```{r}
mean(rivers) # calculate the mean of rivers
```

For the median,

```{r}
median(rivers) # calculate the median of rivers
```

There is no built in function for the mode of a data set. We'll look later at how we can calculate the mode.

But we can calculate the variance and standard deviation,

```{r}
var(rivers) # variance of rivers
sd(rivers) # standard deviation of rivers
```

## Arguments

So far, the only 'argument' we've passed to any of these functions is the data itself.

When calculating the mean, it is not uncommon to trim a percentage from the lower and upper ends of the data set. How much to trim is an argument we can pass to `mean()`. The value we can assign to trim is a fraction from 0 (the default) to 0.5 or 50% of the data set.

```{r}
mean(rivers, trim = 0.1) # drops the upper and lower 10% of the data set from the calculation
```

## Sampling

Another function that is conceptually good to know includes `sample()`, which takes a random sample from a data set. `sample()` can take several arguments, the first of which is a data set to sample, and the second, the number of samples to take.

To take a random sample of 5 values from rivers,

```{r}
sample(rivers, 5) # random sample of 5 from rivers. Will be different every time you run it.
```

This is a useful tool for generating data to test code or building random subsets of a data set to support analysis. We'll explore this application later.

:::note

Other functions to create random samples include `runif()` for uniformly distributed data and `rnorm()` for normally distributed data.

:::

:::summary

Functions are mini programs that do things with our data. They generally have parameters, or arguments, that can be specified to customize how the function operates. 

:::

:::functions

| Function | Description |
| :--- | :--- |
| `mean` | calculate the mean of a range of values. |
| `median` | calculate the median of a range of values. |
| `var` | calculate the variance of a range of values. |
| `sd` | calculate the standard deviation of a range of values. |
| `sample` | take a randmon sample from a range of values. |

:::