---
title: "Importing Data Continued"
output:
  html_document:
    code_folding: show
---

## Dates & Factors

In addition to logical variables, our data set also has categorical data and date data. Categorical data in R is called factor data. We'll convert a single variable to factor to see how this behaves.

```{r}
sq_data$Age <- as.factor(sq_data$Age)

str(sq_data$Age)

levels(sq_data$Age)
```

This process highlights that we have one additional string character that should be converted to `NA` on import, `?`. Calling `str` tells us that we have a factor with three levels or categories. Calling `levels` lists those categories.

Sometime the levels are not in the order we want them for ordinal categorical data. We can re-assign these.

```{r}
sq_data$Age <- factor(sq_data$Age, levels = c("Juvenile", "Adult", "?"))

levels(sq_data$Age)

str(sq_data$Age)
```

<div class = 'exercise'>
Using the help page for `as.Date`, convert the Date variable, currently structured as `int` to a date class variable.
</div>

<div class = 'solution'>
`as.Date` requires a character input, so we must first convert our variable to a character format
```{r, class.source = 'fold-hide', results = FALSE}
sq_data$Date <- as.character(sq_data$Date)
```

Next, we need to specify to `as.Date` how our data is structured - where the year, month, and date are articulated, and whether the year is 4 digits or 2, and the months and days are abbreviated.

```{r, class.source = 'fold-hide', results = FALSE}
sq_data$Date <- as.Date(sq_data$Date, "%m%d%Y")
```
</div>

# Importing with a Data Dictionary

The value of having a data dictionary in advance will go a long ways in facilitating this process.

A data dictionary generally describes key attributes about your data - it will list your variables, provide a description of them, and indicate the base type of data it is.

I have a very simple data dictionary for this project. Let's load that in

```{r}
dict_src <- "https://tinyurl.com/4kxt9n8e" # source of the file

download.file(url = dict_src, destfile = '_datadictionary.csv') # download and name the file

data_dict <- read.csv("_datadictionary.csv") # read in the file
```

We'll now import our data in a slightly more efficient way

```{r}
na_values <- c("" , " ", ".", "NA", "NULL", "?", "+") # a list of values to use for na.strings

data_types <- data_dict$data.import.class # a list of data types to feed to colClass

sq_data <- read.csv(file = "2018-cp_squirrel_census.csv",
         na.strings = na_values,
         colClasses = data_types
           )

str(sq_data)
```

At this point, we've got a pretty clean imported data set. It's likely we'll still find some issues with, but for now let's save it as a cleaned, version 0

```{r}
write.csv(sq_data, "2018-cp_squirrel_census_cleaned_v0.csv", row.names = FALSE)
```

<div class = "summary">
On import, we can control many aspects of our data set, including setting NA values and assigning datat types and classes. This work is facilitated by being organized in advance. Creating a data dictionary will help to facilitate this process.
</div>