---
title: "Loading Data"
output:
  html_document:
    code_folding: show
---

We're going to work with a (hopefully) fun data set today, which we'll download and then import.

First things first, we'll use R to set up a place for us to work. We'll begin by creating a directory on our desktop

```{r create-project-folder, eval = FALSE}
dir.create("~/Desktop/r-for-beginners") # create a project directory
```

<div class = "note">
Most systems - MacOS, Windows, Linux - will set your home directory to your user directory. The tilde `~` is shorthand for your user directory. If your home directory is not your user directory, the above will not work.
</div>

We'll then make sure that our current session is working with the files in this folder.

```{r working-directory, eval = FALSE}
setwd("~/Desktop/r-for-beginners") # set the working directory to the direcotry we just created
```

And then we'll download our data

```{r download-sdata, eval = FALSE}
data_src <- "https://raw.githubusercontent.com/csc-ubc-okanagan/R-for-beginners/main/files/2018-cp_squirrel_census.csv" # define the source of our data, in this case, a url

download.file(url = data_src, destfile = '2018-cp_squirrel_census.csv') # download the file, and save it with a specified name
```

Now that our data is downloaded and we have a local copy, we'll pull it into R

We can simply import the data and display it to our console

```{r import-squirrel-console, eval = FALSE}
read.csv('2018-cp_squirrel_census.csv') # read the data
```

But this isn't terribly useful. We want to store this data in memory, so we'll load it into a variable

```{r import-squirrel-var}
sq_data <- read.csv('2018-cp_squirrel_census.csv') # read the data into a variable called sq_data
```

We can now explore the data, looking at the data types and data structures R is defaulting to when importing this csv file

```{r squirrel-var-str}
str(sq_data) # explore the structure of the data set.
```

We can get slightly better access to the data itself with `View()`

```{r squireel-var-view, eval = FALSE}
View(sq_data) # look at the data in a 'spreadsheet' like format. Note the capital V.
```

This output tells us that we're working with a data frame, that there are 3023 rows or observations, and 31 columns or variables. It then lists all of the variables, tells us what data type they were interpreted as on import, and show us the first six values of each variable.

## Missing Values

Before digging much deeper into the data set, one of the first things we'll note is that there are a lot of missing values. Missing values need to be properly encoded to be programatically useful. An application -- or user -- may represent missing values in many ways. Sometimes it's by using an out of range value, so when a variable is binary, and values are represented as 0 or 1, 99 may be used to indicate a missing value. Files exported from SPSS might be encoded with a period, `.`. In Excel, it's extremely easy for a user to accidentally introduce spaces, ` `, in an otherwise empty cell, or they may choose to write the characters `NA`. A data base export might include either `NA` or `NULL`.

When exported, especially using a format like `csv`, all of these notations for missing values are converted to character strings or numbers. Ideally, when importing into a piece of software, like R, we would have a way to provide a list of possible ways of encoding missing values and standardize how these are presented.

In R, `NA` values are a specific way of indicating that a value is missing. And `read.csv()` has a specific argument for converting strings into `NA` encoded values on import.

<div class = 'exercise'>
Using the help documentation -- `?read.csv` -- see if you can figure out how to update your data import -- `sq_data <- read.csv('2018-cp_squirrel_census.csv')` -- to convert a list of strings to `NA` values.
</div>

<div class = 'solution'>
```{r, class.source = 'fold-hide', results = FALSE}
sq_data <- read.csv('2018-cp_squirrel_census.csv', na.strings = c("", " ", "NA", "NULL", ".", "+"))
```
</div>

```{r, echo = FALSE}
sq_data <- read.csv('2018-cp_squirrel_census.csv', na.strings = c("", " ", "NA", "NULL", ".", "+"))
```

Review the data structure again...

```{r}
str(sq_data)
```

We'll see a number of other things that we need to sort out before we can do much with our data. For example, many of our variable of have not been assigned to the appropriate data type. We have logical variables, date variables etc not yet sorted. Let's fix this.

# Indexing & Data Types

To clean up our data set, we need to be able to target specifc varibales and or specific values. We do this with indexing.

Data frames are comprised of rows and columns, and R accepts arguments in this order, so if we want to see the intersection of the first row and the first column, we would run

```{r}
sq_data[1,1] # show the first row and first column of the data frame
```

If we wanted to see the first 6 rows

```{r}
sq_data[1:6, 1] # show the first six rows and first column of the data frame
```

And if we wanted to see all the values of a specific variable, or column

```{r, eval = FALSE}
sq_data[ ,1]
```

When working with a data frame, we can also use variable names

```{r, eval = FALSE}
sq_data[1:6, "Running"]
```

And when we want every observation associated with a given variable

```{r, eval = FALSE}
sq_data$Running
```

## Changing Data Types

Now that we can target a variable, we can assign new properties to that variable. R has a series of `as.` functions to allow us to manipulate data types and classes. To view all `as.` functions available to you

```{r, eval = FALSE}
apropos("^as\\.")
```

```{r, echo = FALSE}
head(apropos("^as\\."))
cat("...")
```

We can see in the list, that we have several useful `as.` functions for our current data set, including `as.Date` and `as.logical`. We'll start with the latter.

## Convert a single variable

```{r}
sq_data$Running <- as.logical(sq_data$Running) # re-assign the variable Running from character to logical

str(sq_data$Running) # view the results
```

## Convert multiple variables

```{r}
variables_to_convert <- c("Running", "Chasing")

sq_data[,variables_to_convert] <- lapply(sq_data[,variables_to_convert], as.logical)

str(sq_data$Chasing)
```

This is a steep introduction to vectorization in R.

Remember that a vector is just a simple list. Generally, when we do one thing to something in a list it only impacts that one value -- in Excel we might double one value in a column, but this doesn't double every value in the column.

R is built from the ground up to apply changes to all values in a list -- known as vectorization. So, if we have a series of numbers -- like we created on earlier with `random_sample`

```{r}
random_sample
```

And we want to convert every value, we apply the conversion to the entire vector

```{r}
(random_sample_doubled <- random_sample * 2)
```

The `apply` family of functions does exactly the same thing, but instead, in the case of `lapply`, taking in a single object, like `random_sample`, it takes in a list of objects -- hence the `l` for `list apply` --  and applies a function to each object.

Reading out the above chunk of code then

```{r, eval = FALSE}
# create a character vector with the names of our variables
variables_to_convert <- c("Running", "Chasing") 

## create a list of variables - sq_data[, "Running"], sq_data[, "Chasing"] - that is fed
## into the lapply() function and apply the as.logical() function to each item
sq_data[,variables_to_convert] <- lapply(sq_data[,variables_to_convert], as.logical)
```

Writing code in this way can be extremely helpful, starting with a plain language write up and then translating each piece to actual code:

1. write a list of variables we want to change to a variable
2. index the data frame with the above variable
3. apply a function to this index

Writing it like this also allows you to articulate your problem in a language that can be searched if you don't know how to complete a specific step.

<div class = "advanced">
Without vectorization, you would need to specify how to iterate over each value in an object. Sometimes you still need to do this in R, so it might be handy to see how this works. Working with a plain language summary and using our `random_sample`:

1. create an object to hold our output - call it `random_sample_multi`
2. start an iteration that begins at index point 1 - `random_sample[1]` - and ends at the final index point of the object we're working with - `random_sample[20]`, remembering that writing `random_sample[1]` will output the value held at that index point and that we don't need to know the final index point, as we can calculate this with `length()`.
3. populate the object we created in step one with the multiple of 2 of each value held in each index point of `random_sample`
4. print out the result to confirm

```{r}
random_sample_multi <- vector() # create an empty vector called random_sample_multi

for (i in 1:length(random_sample)) { # iterate with a for loop starting with the number 1 and ending with the last index point in random sample, calculated here using length()
  random_sample_multi[i] <- random_sample[i] *2 # for each value at index point 1 through to length(random_sample), multiply that value by 2
}

random_sample_multi # print the result to verify
```
</div>

<div class = 'exercise'>
`read.csv` allows us to assign variable types on import. Revisit the help documentation for `read.csv` and see if you can update your import line -- `sq_data <- read.csv('2018-cp_squirrel_census.csv', na.strings = c("", " ", "NA", "NULL", ".", "+"))` -- to also convert the columns 'Running', 'Chasing', and 'Climbing' to logical on import.
</div>

<div class = 'solution'>
```{r, class.source = 'fold-hide', results = FALSE}
sq_data <- read.csv('2018-cp_squirrel_census.csv',
                    na.strings = c("", " ", "NA", "NULL", ".", "+"),
                    colClasses = c("Chasing" = "logical",
                                   "Running" = "logical",
                                   "Climbing" = "logical")
                    )
```
</div>

## Dates & Factors

In addition to logical variables, our data set also has categorical data and date data. Categorical data in R is called factor data. We'll convert a single variable to factor to see how this behaves.

```{r}
sq_data$Age <- as.factor(sq_data$Age)

str(sq_data$Age)

levels(sq_data$Age)
```

This process highlights that we have one additional string character that should be converted to `NA` on import, `?`. Calling `str` tells us that we have a factor with three levels or categories. Calling `levels` lists those categories.

Sometime the levels are not in the order we want them for ordinal categorical data. We can re-assign these.

```{r}
sq_data$Age <- factor(sq_data$Age, levels = c("Juvenile", "Adult", "?"))

levels(sq_data$Age)

str(sq_data$Age)
```

<div class = 'exercise'>
Using the help page for `as.Date`, convert the Date variable, currently structured as `int` to a date class variable.
</div>

<div class = 'solution'>
`as.Date` requires a character input, so we must first convert our variable to a character format
```{r, class.source = 'fold-hide', results = FALSE}
sq_data$Date <- as.character(sq_data$Date)
```

Next, we need to specify to `as.Date` how our data is structured - where the year, month, and date are articulated, and whether the year is 4 digits or 2, and the months and days are abbreviated.

```{r, class.source = 'fold-hide', results = FALSE}
sq_data$Date <- as.Date(sq_data$Date, "%m%d%Y")
```
</div>

# Importing with a Data Dictionary

The value of having a data dictionary in advance will go a long ways in facilitating this process.

A data dictionary generally describes key attributes about your data - it will list your variables, provide a description of them, and indicate the base type of data it is.

I have a very simple data dictionary for this project. Let's load that in

```{r}
dict_src <- "https://raw.githubusercontent.com/csc-ubc-okanagan/R-for-beginners/main/files/_datadictionary.csv" # source of the file

download.file(url = dict_src, destfile = '_datadictionary.csv') # download and name the file

data_dict <- read.csv("_datadictionary.csv") # read in the file
```

We'll now import our data in a slightly more efficient way

```{r}
na_values <- c("" , " ", ".", "NA", "NULL", "?", "+") # a list of values to use for na.strings

data_types <- data_dict$data.import.class # a list of data types to feed to colClass

sq_data <- read.csv(file = "2018-cp_squirrel_census.csv",
         na.strings = na_values,
         colClasses = data_types
           )

str(sq_data)
```

At this point, we've got a pretty clean imported data set. It's likely we'll still find some issues with, but for now let's save it as a cleaned, version 0

```{r}
write.csv(sq_data, "2018-cp_squirrel_census_cleaned_v0.csv", row.names = FALSE)
```

<div class = "summary">
On import, we can control many aspects of our data set, including setting NA values and assigning datat types and classes. This work is facilitated by being organized in advance. Creating a data dictionary will help to facilitate this process.
</div>

<div class = 'functions'>
| Function | Description |
| :--- | :--- |
| `dir.create` | create a directory on your file system. |
| `setwd` | set the working directory. See also `getwd` to get current working directory. |
| `read.cs` | read a csv file into R. |
| `str` | display information about the data including structure, types, and a few values. |
| `View` | open the data set in a spreadsheet like viewer. |
| `apropos` | search for functions and variables. |
| `as.` | a family of functions for converting data types and structures. |
| `lapply` | apply a function to every item in a list. |
| `levels` | print the categories and codes for categorical data. |
| `write.csv` | export data from R to your file system in csv format. |
</div>