---
title: ggplot2
output:
  html_document:
    code_folding: show
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE
)

library(ggplot2)
library(palmerpenguins)
```

<div class = "note">
This is a brief introduction. For more in depth examples and solutions, check out [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html) by Hadley Wickham, Danielle Navarro, and Thomas Lin Pedersen.
</div>

`ggplot()` is based on a grammar graphics -- a way of approaching describing the construction of a graphic from common building blocks. Building a graphic with ggplot then follows some common patterns of construction.

At its most basic, we supply ggplot with a data set and some aesthetics to map -- that is, the variables we wish to display on the plot and how they should appear. We then define a plot type.

Step by step this looks like

1. call ggplot()
2. provide ggplot with a data set
3. provide ggplot with the variables of interest and their aesthetic properties
4. define a plot type with geom_*plotType*()

After this, we continue to add on layers. We can add on additional `geoms`, we can add on scale adjustments, statistical summaries etc. and finally we can adjust how the overall presentation looks, adjusting titles, labels, legends etc. The general form of this is something like:

```{r demo_ggplot, eval = FALSE}
ggplot() + # call ggplot and potentially feed in data and variables to plot to specific aesthetics
  geom() + # pick a plot type. Data and variables to plot to specific aesthetics can also be assigned here
  scale() + # make adjustments to the scales
  labs() + # customize labels
  ... + # there are other options!
  theme() # add some style
```

## Levels of Measurement & Data Types

We should briefly discuss levels of measurement. A common taxonomy breaks values into four levels:

| Level | Order | Description | Example | General Note |
| :---| :--- | :--- | :--- | :--- |
| Nominal | N | Classifies | Marital status | Pick a category |
| Ordinal | Y | Classifies, > < comparisons | Education | Pick a preference |
| Interval | Y | Difference, - + comparisons | Number of people | Count something |
| Ratio | Y | Magnitude, x / comparisons | Height | Measure something |

Not all visuals are appropriate for all levels. As a consequence, the options available to the various `geoms` that define the plot type will differ.

Reflecting back on our earlier discussion of data types in R, nominal and ordinal data are generally categorical or 'factor', interval will be classified as integers, and ratio will be classified as double. If our data types are not properly assigned, R will generally try to 'coerce' the data type to something that will work, but this isn't always successful, nor is very good practice to make the system guess!

## Get Some Data

First we get some data

```{r install_pp, eval = FALSE}
install.packages("palmerpenguins")

library(palmerpenguins)
```

And of course, check out the data

```{r str_pp}
str(penguins)
```

<div class = "note">
The penguins data structure is a `tibble`. A tibble is very much like a data frame. It is not part of base R, but is introduced through Tidyverse, of which ggplot2 is one piece.
</div>

And optionally, view the data

```{r view_pp, eval = FALSE}
View(penguins)
```

## The Basic Graph

Following on the above, first we call ggplot and define our data set:

```{r ggplot_data}
ggplot(data = penguins)
```

An blank slate is prepared for us! Next, we add in the variables we want mapped.

```{r ggplot_aes}
ggplot(data = penguins, aes(x = bill_length_mm, y = flipper_length_mm))
```

This sets us up with a blank grid with x and y axis ticks corresponding to your variable values and x and y axis labels corresponding to your variables.

Now we call a plot type to represent our data, in this case, a scatter plot

```{r ggplot_point}
ggplot(data = penguins, aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point()
```

There are several `geoms_` that we can choose from. These include:

* geom_bar() and geom_col() for bar charts
* geom_boxplot() for box and whisker plots
* geom_density() and geom_violin() for density plots
* geom_freqpoly(), geom_histogram(), and geom_dotplot for distributions
* geom_line() for connecting observations
* geom_point() and geom_jitter() for scatter plots
* geom_map() and geom_sf() for GIS data

There are more...

<div class = "note">
There are many aesthetics that we can map to `aes()` and each `geom_` will accept different aesthetics. You can review available aesthetics using the help pages. Those aesthetics in **bold** are required.
</div>

<div class = "exercise">
Using the above as a model, use `geom_jitter()` to plot `flipper_length_mm` against `species`.
</div>

<div class = "solution">
```{r solution_1, class.source = 'fold-hide'}
ggplot(data = penguins, aes(x = species, y = flipper_length_mm)) +
  geom_jitter()
```
</div>

## A Basic Bar Graph

We'll now plot a bar plot, which only maps one variable as the other variable is simply a count of observations, which is computed as part of the process of generating the graph.

```{r basic_bar}
ggplot(data = penguins, aes(x = sex)) +
  geom_bar()
```

If we already had the count data, we would use `geom_col()`, but then we need to provide the y aesthetic. R provides an easy way to get a frequency table with the function `table()`:

```{r freq_table}
(sex_freqtable <- table(penguins$sex))
```

`NA` values are dropped by default.

We can't plot a table. To be able to plot this, we need to convert the table into a data frame.

<div class = "exercise">
Use `geom_col()` to plot `sex_freqtable`. A plain language articulation of the process would read something like:

1. convert `sex_freqtable` to a dataframe
2. figure out the variable names in the dataframe that is created
3. call ggplot, feed it the data frame and the variables to be plotted
4. add `geom_col()`
</div>

<div class = "solution">
```{r solution_2, class.source = 'fold-hide', results = FALSE}
sex_freqtable_df <- as.data.frame(sex_freqtable) # convert to data frame

str(sex_freqtable_df) # pull up the struture

ggplot(data = sex_freqtable_df, aes(x = Var1, y = Freq)) +
  geom_col()
```
</div>

## Additional Aesthetics

Beyond articulating the variables on the x and y plane, we can also articulate variables according to other aesthetics properties, such as colour and size.

Using a bar plot

```{r bar_fill}
ggplot(data = penguins, aes(x = sex, fill = species)) +
  geom_bar()
```

Let's get rid of those `NA` values. ggplot allows us to subset.

```{r bar_fill_na}
ggplot(data = subset(penguins, !is.na(sex)), aes(x = sex, fill = species)) +
  geom_bar()
```

Using our jitter plot

```{r jitter}
ggplot(data = penguins, aes(x = species, y = flipper_length_mm, colour = island)) +
  geom_jitter()
```

<div class = "exercise">
Using the penguins data set, use `geom_point()` to plot `flipper_length_mm` against `bill_length_mm` only for those instances where the variable `sex` is not `NA`, and map the aesthetic `size` to the variable `sex`.

A plain language summary might read:

1. call ggplot
2. feed it the penguins data set subsetted to remove `NA` values from the variable `sex`
3. map the variable `bill_length_mm` to the x axis, `flipper_length_mm` to the y axis, and `sex` to size
4. call `geom_point()`
</div>

<div class = "solution">
```{r size_point, class.source = 'fold-hide'}
ggplot(data = subset(penguins, !is.na(sex)), aes(x = bill_length_mm, y = flipper_length_mm, size = sex)) +
  geom_point()
```
</div>

<div class = "note">
You should get a warning that reads `Using size for a discrete variable is not advised.` This is valid. Size comparisons are one of magnitude while the `sex` variable represents two discrete, non-ordinal categories. The graph can be produced, but that doesn't mean it's a good graph for the data.
</div>

It would be more appropriate to use colour to distinguish between a non-ordinal, categorical variable and to use size for something like body mass.

```{r colour_point}
ggplot(data = subset(penguins,!is.na(sex)),
       aes(x = bill_length_mm,
           y = flipper_length_mm,
           colour = sex,
           size = body_mass_g)) +
  geom_point()
```

Things get a little hard to see here though!

<div class = "note">
To keep things legible in our code, you may wish to use a line break at the end of each argument.
</div>

## Scales

Our scales can be adjusted in many different ways, for example, changing the colour, fill, alpha, x, or y scales.

<div class = "note">
`alpha` controls opacity and is on a scale of 0 to 1, with 0 being transparent and 1 being solid. Like jitter, this helps us to see overlapping data points.
</div>

We'll start by mapping the variable `sex` to the `alpha` aesthetic and then adjust the alpha scale, which allows us to set each category at a different opacity. 

```{r alpha_point}
ggplot(subset(penguins, !is.na(sex)),
       aes(x = bill_length_mm,
           y = flipper_length_mm,
           colour = sex,
           size = body_mass_g,
           alpha = sex)) +
  geom_point() +
  scale_alpha_manual(values = c(0.3, 0.5)) # we have to levels to our factor, so we need to supply two values 
```

Since alpha only refers to colour, we can also provide the argument with a single value within the aesthetic mapping, which maps that alpha value to all variables

```{r alpha_point_2}
ggplot(subset(penguins, !is.na(sex)),
       aes(x = bill_length_mm,
           y = flipper_length_mm,
           colour = sex,
           size = body_mass_g,
           alpha = 0.5)) +
  geom_point()
```

The result is a little odd, since 0.5 is now being treated as a variable that requires a legend. We'll sort that one out later!

In these cases, we've manually adjusted the alpha scale. Similarly, we could manually adjust the colour scale

```{r colour_point_alpha}
ggplot(subset(penguins, !is.na(sex)),
       aes(x = bill_length_mm,
           y = flipper_length_mm,
           colour = sex,
           size = body_mass_g,
           alpha = 0.5)) +
  geom_point() +
  scale_colour_manual(values = c("blue", "green"))
```

We don't need to complete all scales manually. One situation in which this might be handy is for colours. There is a great tool called ColourBrewer that has set scales for working with non-ordinal, ordinal, and ratio data.

Install the package and load the library

```{r install_cb, eval = FALSE}
install.packages("RColourBrewer")
```

```{r load_cb}
library(RColorBrewer)
```

Check out what's available:

```{r list_cb, fig.height= 8}
display.brewer.all()
```

Pick a colour set and add the scale

```{r custom_cb}
ggplot(subset(penguins, !is.na(sex)),
       aes(x = bill_length_mm,
           y = flipper_length_mm,
           colour = sex,
           size = body_mass_g,
           alpha = 0.5)) +
  geom_point() +
  scale_colour_brewer(palette = "Set1")
```

## Labeling and captions

Labels default to our variable names, which may not be what we want on our graph. We can override this with `labs()`. We'll investigate this with a box plot.

```{r labs_1}
ggplot(data = subset(penguins, !is.na(sex)), aes(x = sex, y = body_mass_g)) +
  geom_boxplot() +
  labs(
    x = "Sex",
    y = "Body Mass (grams)"
  )
```

We can also add titles and captions.

```{r labs_2}
ggplot(subset(penguins, !is.na(sex)), aes(x = sex, y = body_mass_g)) +
  geom_boxplot() +
  labs(
    x = "Sex",
    y = "Body Mass (grams)",
    title = "Palmer Penguins",
    caption = "Palmer Penguins. Body mass grouped by sex. n = 333."
  )
```

And our text can leverage variables, using for example `paste0()`, which combines strings together.

```{r}
observations <- nrow(subset(penguins, !is.na(sex))) # count the number of observations

ggplot(subset(penguins, !is.na(sex)), aes(x = sex, y = body_mass_g)) +
  geom_boxplot() +
  labs(
    x = "Sex",
    y = "Body Mass (grams)",
    title = "Palmer Penguins",
    caption = paste0("Palmer Penguins. Body mass grouped by sex. n = ", observations, ".") # print the number of observations
  )
```

## More than one geom

More than one geom can help to convey more information. To do this, we feed our data set and aesthetic mappings into `ggplot()`, and then call multiple geoms to represent those variables.

```{r}
ggplot(subset(penguins, !is.na(sex)), aes(x = sex, y = body_mass_g)) +
  geom_boxplot() +
  geom_jitter(width = 0.10) + # control the width of the jitter geom
  labs(
    x = "Sex",
    y = "Body Mass (grams)",
    title = "Palmer Penguins",
    caption = paste0("Palmer Penguins. Body mass grouped by sex. n = ", observations, "."))
```

## More than one plot

There are several ways to place more than one plot side by side. One of the easiest is to use patchwork.

Install

```{r, eval = FALSE}
install.packages("patchwork")
```

Load

```{r}
library(patchwork)
```

We can store plots inside variables to be called when we want. This is necessary with patchwork.

```{r}
base_plot <- ggplot(subset(penguins, !is.na(sex)), aes(x = sex, y = body_mass_g))

penguins_boxplot <- base_plot +
  geom_boxplot()

penguins_jitter <- base_plot +
  geom_jitter(width = 0.10)
```

We can then use patchwork to arrange our plots

```{r}
penguins_boxplot + penguins_jitter
```

```{r}
penguins_boxplot / penguins_jitter
```

<div class = "note">
There are many ways in which patchwork can arrange plots. See the chapter [Arranging Plots](https://ggplot2-book.org/arranging-plots.html) in *ggplot2: Elegant Graphics for Data Analysis* for more complex examples.
</div>

## Faceting a Plot

ggplot also has built in faceting options, `facet_grid()` and `facet_wrap()`, allowing use to break a plot on discrete variables.

```{r facet_grid_1}
penguins_jitter +
  facet_grid(cols = vars(island))
```

`facet_grid()` produces produces a grid for each combination of variables, whether there are values for that combination or not.

```{r facet_grid_2}
penguins_jitter +
  facet_grid(cols = vars(species), rows = vars(island))
```

`facet_wrap()` on the other hand, only produces a grid for those variables with combinations.

```{r facet_grid_3}
penguins_jitter +
  facet_wrap(vars(species, island))
```

## Themes {-}

There are many built in themes, as well as ways to customize themes.

```{r density_1}
ggplot(penguins, aes(x = flipper_length_mm, fill = species, colour = species)) +
  geom_density(alpha = 0.3) +
  theme_minimal()
```

Within a theme, we can start to customize other elements. Things that we can customize included axes elements, legend elements, panel elements, and plot elements. For example, we can build on the theme minimal and remove the panel grids above, we do this with a separate, additional call to theme():

```{r density_2}
ggplot(penguins, aes(x = flipper_length_mm, fill = species, colour = species)) +
  geom_density(alpha = 0.3) +
  theme_minimal() +
  theme(
    panel.grid = element_blank()
  )
```

<div class = "note">
A full list of theme options are available on the [ggplot theme reference page](https://ggplot2.tidyverse.org/reference/theme.html).
</div>