---
title: "Variables"
output:
  html_document:
    code_folding: show
---

```{r echo=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      comment = NA)
```

When we have a function do something with our data, there are several ways in which we might direct the output of that function. So far, we've directed that output to our terminal -- or screens. We have other options though. We can direct that output to a file if needed. We can also store the value within what R calls our global environment, which for all intents and purposes means writing the output to memory. We do this by directing that output to, or storing that output in, a variable.

:::note

When we collect data, we assign these to a variable, like 'ht' to record height. When we need to store values in memory on our computer, we assign them a variable, or name, by which to update or recall them. In this way, we can store, for example, the output of a calculation, such as the output of `mean(rivers)` with reference to a name.

Sometimes when we say 'variable' we'll mean the category of data you've collected and sometimes we'll mean a place holder for a stored value on your computer.

:::

In R, things are assigned to variables with the lesser than symbol `<` and a hyphen `-`, like `<-`, as in, put what's on the right of this into the variable on the left.

To store the mean of rivers in a variable called `rivers_mean`,

```{r}
rivers_mean <- mean(rivers) # store the value of mean(rivers) in the variable rivers_mean
```

And to recall that value,

```{r}
rivers_mean # print the value stored in the variable rivers_mean
```

or run a computation with that value

```{r}
rivers_mean ^ 2 # square the mean
```

:::note

When naming variables in `R`, keep in mind that variable names:

* Should first and foremost be meaningful. This is not a rule, just best practice.
* **Cannot** start with a number or a dot followed by a number.
* **Cannot** contain spaces or hyphens.
* **Can** contain letters, numbers, dots, and underscores.

Additionally, some words are *reserved* and cannot be used, such as *if* and *for*. More details can be found with `?make.names`

:::   

:::exercise

Save the output of a random sample of 10 digits between 1 and 100 to a variable called `random_sample`.

:::

:::solution

```{r, class.source = 'fold-hide'}
(random_sample <- sample(1:100, 10))
```

:::