---
title: Subsetting
output:
  html_document:
    code_folding: show
---

There are three subsetting operators in R -- [[, [, $ -- each of which will work with a data frame.

We've already seen how we can select a named column in a data frame with the dollar sign selector.

```{r, eval = FALSE}
sq_data$Age
```

```{r, echo = FALSE}
head(sq_data$Age)
```

We can also use indexed positions with square brackets

```{r, eval = FALSE}
sq_data[8]
```

```{r, echo = FALSE}
head(sq_data[8])
```

If we only include a single digit, R defaults to selecting a column.

We can select both rows and columns, however, using a comma, where the first number, or set of numbers, selects the rows, and the second, the columns.

```{r}
sq_data[1,1] # show the first row and first column of the data frame
```

If we wanted to see the first 6 rows

```{r}
sq_data[1:6, 1] # show the first six rows and first column of the data frame
```

Leaving a selection blank is the equivalent of selecting everything; either all rows, or all columns,

```{r, eval = FALSE}
sq_data[ ,1] # all rows, first column, ie one variable
sq_data[1, ] # all columns, first row, ie one observation
```

We can also mix and match names and indexes,

```{r, eval = FALSE}
sq_data[1:6, "Running"]
```

:::exercise

There are two ways to select a single column in a data frame; using [] or [ , ]. Extract a single variable, and all observations, from `sq_data` using each of these two methods, saving each to their own variable. Then using `str()` investigate the difference in the outputs.

:::

:::solution

```{r, class.source = 'fold-hide'}
matrix_ss <- sq_data[6]
str(matrix_ss)

list_ss <- sq_data[ ,6]
str(list_ss)
```

:::

:::exercise

We've encountered `sample()`. We can use sample and index subsetting take a random sample from our data set. To do this, we'll need to use the `nrow()` function, that when fed a data set, reports on the number of rows,

```{r}
nrow(sq_data)
```

Using this information, Take a random sample of 10 observations for `sq_data` for the first 6 columns.

:::

:::solution

```{r, class.source = 'fold-hide'}
sq_data[sample(nrow(sq_data), 10), 1:6]
```

:::


Lastly, we might want only observations that are complete, that is, records with no missing values

```{r}
(sq_data_complete <- sq_data[complete.cases(sq_data), ]) # not terribly useful with this particular data set
```

### Logical subsetting

So far, we haven't applied any conditions to our subsets, but we can. In logical subsetting, the subset defaults to returning the results where the condition is TRUE.

R allows us to specify several conditions

* less than `>`
* greater than `<`
* less than or equal to `<=` 
* greater than or equal `>=`
* equivalent to `==`
* not equivalent to `!=`

As well as boolean operators

* or `|`
* and `&`

:::note

Note that `=` is equivalent to `<-`, setting a value, while `==` tests whether or not two things are the same.

```{r}
(x = 2)
(x == 2)
```

:::

```{r, eval = FALSE}
sq_data[sq_data$Age == "Adult", ] # all variables for all Adult observations

sq_data[sq_data$Age == "Adult" & sq_data$Shift == "PM", ] # all variables for all Adult, PM observat
```

### Subset()

A lot this functionality has been built into a subset function, `subset()`.

We'll provide `subset()` with three arguments:

1. A data set to subset from
2. The condition on which to subset
3. The variables from the original data set to keep -- leave this empty to default to all original variables

Let's say we only want to see the records for adult squirrels.

```{r}
sq_data_adult <- subset(sq_data, Age == "Adult") # from sq_data, select only rows where Age equals Adult
```

Or all records with cinnamon coloured adult squirrels.

```{r}
sq_data_c_adult <- subset(sq_data, Age == "Adult" & Primary.Fur.Color == "Cinnamon")
```

Or all records where an age has been recorded. This can be done in two ways

```{r}
sq_data_age_1 <- subset(sq_data, !is.na(Age))

sq_data_age_2 <- subset(sq_data, Age == "Adult" | Age == "Juvenile")
```

:::note

Remember, `NA` values are a special value, indicating that a value is missing. This is different from writing the characters NA. `is.na()` tests for `NA`. It's for this reason that running `subset(sq_data, sq_data$Age == 'NA')` will give you an empty data frame.

:::

Or maybe we want to investigate patterns of only those squirrels for which we don't have age recorded!

```{r}
sq_data_age_na <- subset(sq_data, is.na(Age))
```

:::exercise

All data was collected between October 6 and October 20, 2018.  Create a subset of the data that includes only data collected on or before October 14, and only retains the variables `Date` and `Age`. Verify your date range output with `unique()` and if you get that far, order this output from `unique()` using `sort()` for easier reading.

Remember that we have not yet converted our `Date` variable to a date class. 

:::

:::solution

```{r, class.source = 'fold-hide', results = FALSE}
sq_data$Date <- as.Date(sq_data$Date, "%m%d%Y")
sq_data_date <- subset(sq_data, Date <= "2018-10-14", select = c(Date, Age))
sort(unique(sq_data_date$Date))
```

:::

:::summary

Subsetting can be achieved in many ways. We can use indexing to get specific observations (rows) and / or variable (columns), we can use conditions on which to subset, we can use other functions like `is.na()` and `complete.cases()` to work around missing values, and we can use `subset()`, among other functions, subset.

:::

:::functions

| Function | Description |
| :--- | :--- |
| `subset` | conditionally susbet a data frame |
| `is.na` | conditionally test if values in a vector are `NA`. |
| `complete.cases` | conditionally test if cases (rows) are complete. |

:::