```{r variables-storing}
 # generate a random sample of 20 data points and store it in a variable called random_sample
random_sample <- sample(1:100, 20, replace =TRUE)

# call random_sample which outputs the values to the screen
random_sample

# calculate the mean of random_sample and store it in a variable called random_sample_mean
random_sample_mean <- mean(random_sample) 

# call random_sample_mean which outputs the values to the screen
random_sample_mean 
```

<div class = 'exercise'>
Let's build our own function that calculates a mean to see what's all happening under the hood.

First we define the mean as the sum of all data points divided by the number of data points. We'll leverage two existing functions, `sum()` and `length()`, the former of which performs simple addition and the latter of which tells us how many data points are in a vector.
</div>

<div class = 'solution'>
```{r mean-function, class.source = 'fold-hide', results = FALSE}
# define our function. The function takes one input, which is a data source, it then add all the data points together and divides it by the number of data points.
# the function is stored in a variable called my_mean
my_mean <- function(data) {
  total <- sum(data) # add the data
  n = length(data) # count the data
  calculation = total/n # do the math
  calculation # print the output
}

my_data <- sample(1:100, 20, replace = F) # generate some data

my_mean(my_data) # calculate the mean using my_mean
```
</div>

<div class = 'summary'>
Functions take in data and do things with that data. Variables allow us to store our data for easy access.
</div>


## Convert multiple variables

We might be inclined to simply do the following:

```{r, error = TRUE}
variables_to_convert <- c("Running", "Chasing")

sq_data[,variables_to_convert] <- as.logical(sq_data[,variables_to_convert])
```

Which will give us an error. Errors in R are not always super helpful and working through them is a learning curve unto itself. This one is somewhat helpful. It tells us that as.logical doesn't know how to handle a list as an input. A plain language interpretation of this is that `as.logical()` will only take in a single vector of values, not more than one vector.

The solution is to use the `apply()` family of functions.

```{r}
variables_to_convert <- c("Running", "Chasing")

sq_data[,variables_to_convert] <- lapply(sq_data[,variables_to_convert], as.logical)

str(sq_data$Chasing)
```

The `apply` family of functions is very useful for iterating over data structures, such as data frames. In `lapply()`, the l stands for list, suggesting it takes in lists to iterate or loop over. In this case, the function takes a list, `variables_to_convert`, and passes that list off, one vector at a time to `as.logical()`.

<div class = "advanced">
We're about to take a somewhat steep introduction to two key concepts in R related to iterating operations over your data in R.

There are several ways that one might iterate operations. One of the most common is a for loop, in which we state, for each element in a list, do this to it. This would look something like this, using our earlier `random_sample` we created:

```{r}
random_sample_doubled <- vector() # create an empty vector called random_sample_multi

for (i in 1:length(random_sample)) { # iterate with a for loop starting with the number 1 and ending with the last index point in random sample, calculated here using length()
  random_sample_doubled[i] <- random_sample[i] *2 # for each value at index point 1 through to length(random_sample), multiply that value by 2
}

random_sample_doubled # print the result to verify
```

R is a high level language, meaning it's interpreted at run time rather than pre-compiled. This generally makes for loops slow, since R takes time to process each individual step; do something to the first element, then to the second element etc. There are ways to make for loops faster in R; one way is simply to let the apply family of functions take care of the details for us. We don't need to write code that explicitly loops. Instead, we call an apply function that is designed to efficiently handle the task for us.

The second key concept here is to leverage when possible the fact that R is vectorized. We see this with examples like this where we don't need to loop over every value, its the default behaviour.

```{r}
random_sample_2 <- random_sample * 2
```

What this means 'under the hood' though, is that instead of R locally processing your data, slowly iterating over it, R looks at the first element of your vector, determines its data type, and then passes the operation off to a function written in a pre-compiled language. Pre-compiled languages are much faster. If you dive into R more deeply, you'll hear people say that you should vectorize your code. What this really suggests is, if you find yourself writing a lot of iterative loops, like for loops, consider that there might be a more efficient path that involves passing vectors to already pre-compiled functions.
</div>
